---
title: "Benchmarking fastINFLECT against consensus metaclustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking fastINFLECT against consensus metaclustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)

# Precomputed by data-raw/make-benchmark-cache.R. ConsensusClusterPlus is slow,
# and the original-INFLECT baseline is projected from sampled legacy dip.test
# cells, so timings are measured once offline and displayed here.
local_cache_candidates <- c(
  file.path("..", "inst", "extdata", "benchmark_cache.rds"),
  file.path("inst", "extdata", "benchmark_cache.rds")
)
local_cache <- local_cache_candidates[file.exists(local_cache_candidates)]
cache_path <- if (length(local_cache)) {
  local_cache[[1L]]
} else {
  system.file("extdata", "benchmark_cache.rds", package = "fastINFLECT")
}
cache <- readRDS(cache_path)
library(ggplot2)

fmt_secs <- function(x) ifelse(x < 1, sprintf("%.2f s", x), sprintf("%.1f s", x))
fmt_x <- function(x) sprintf("%.0f×", x)
pkg_version <- function(pkg) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    as.character(utils::packageVersion(pkg))
  } else {
    NA_character_
  }
}
vignette_reproducibility <- function() {
  data.frame(
    Component = c("R", "fastINFLECT", "FlowSOM", "diptest"),
    Version = c(
      as.character(getRversion()),
      pkg_version("fastINFLECT"),
      pkg_version("FlowSOM"),
      pkg_version("diptest")
    ),
    stringsAsFactors = FALSE
  )
}
```

## The task being benchmarked

Choosing the number of metaclusters *k* for a FlowSOM/SOM requires **scanning a
range of k and judging cluster quality at each**. This vignette benchmarks three
ways of doing that on the same SOM:

| Approach | How it scans k | How it chooses k |
|---|---|---|
| **Original INFLECT** | scores each k independently with the original `FlowSOMQC()` / `dip.test()` loop | automatic inflection |
| **FlowSOM consensus** | run `metaClustering_consensus()` (ConsensusClusterPlus) once per k | left to the user |
| **fastINFLECT** | one memoised sweep that cuts the SOM's hierarchy and scores marker unimodality | automatic (inflection / kneedle / unimodality threshold) |

The data is the bundled downsampled Levine *et al.* (2015) CyTOF dataset: a
FlowSOM object with **`r cache$machine$cores`-core** timings on
`r cache$machine$sysname`, `r cache$machine$r_version`, fastINFLECT
`r cache$machine$inflect_version`. Absolute times are machine-specific; the
*ratios* are the point.

> The methods are not doing the same statistics — consensus clustering builds
> a stability-based partition, while fastINFLECT cuts a Ward hierarchy and scores marker
> unimodality. The benchmark compares the **end-to-end cost of scanning k to build a
> quality curve**, which is the practical bottleneck, and highlights that fastINFLECT
> additionally returns a recommended k.

---

## The scan patterns

```{r inflect-scan, eval = FALSE}
library(fastINFLECT)
library(FlowSOM)
load(system.file("extdata", "Levine32sample.Rdata", package = "fastINFLECT"))

# fastINFLECT: one call scores every k and recommends one.
inflect_res <- INFLECT(dataset, set.i = 5:25, multicore = FALSE, zeroes.in = FALSE)
```

```{r original-inflect-scan, eval = FALSE}
# Original INFLECT pattern: one FlowSOMQC-style score per k, using dip.test()
# inside every cluster-marker cell. The benchmark projects this cost from sampled
# original cells because a full old-engine run is intentionally slow.
ml <- iteration.metacluster(dataset, set.i = 5:25, multicore = FALSE)
legacy_qc <- lapply(5:25, function(k) {
  FlowSOMQC(dataset, ml[[as.character(k)]], zeroes.in = FALSE, verbose = FALSE)
})
```

```{r consensus-scan, eval = FALSE}
codes <- dataset$map$codes

# (a) Convenience pattern: one metaClustering_consensus() call per k. Each call
#     runs a fresh ConsensusClusterPlus up to maxK = k, so work is repeated.
for (k in 5:25) mc_k <- metaClustering_consensus(codes, k = k, seed = 42)

# (b) Amortized: a single ConsensusClusterPlus run to maxK yields every k at once
#     (this is what (a) slices internally). The fairest consensus baseline.
ccp <- ConsensusClusterPlus::ConsensusClusterPlus(t(codes), maxK = 25, seed = 42,
                                                  plot = NULL, verbose = FALSE)
# ... score ccp[[k]]$consensusClass for each k ...
```

---

## Runtime

Time to scan `k = 5:25` (21 values) and obtain a quality score at every k. Both
consensus variants produce identical partitions; only their cost differs. The
original-INFLECT bar is a conservative projection from
`r cache$efficiency$legacy_projection$sampled_tests` sampled original
`dip.test()` cluster-marker cells and excludes small metaclustering and curve-fit
overhead.

```{r runtime-bar, fig.height = 3.6, fig.cap = "Wall-clock to scan k = 5:25. fastINFLECT evaluates every distinct SOM-node subtree once and reuses it across all k. Original INFLECT is projected from sampled original dip.test cells. A single ConsensusClusterPlus run is the fair consensus baseline; calling metaClustering_consensus() per k repeats that work 21 times."}
rt <- data.frame(
  method = c("fastINFLECT 1.0",
             "Original INFLECT\n(projected)",
             "FlowSOM consensus\n(single run)",
             "FlowSOM consensus\n(per-k convenience)"),
  seconds = c(cache$totals$inflect_scan_seconds,
              cache$totals$legacy_inflect_scan_seconds,
              cache$totals$amortized_seconds,
              cache$totals$consensus_scan_seconds)
)
rt$method <- factor(rt$method, levels = rt$method[order(rt$seconds)])
rt$engine <- c("fastINFLECT 1.0", "Original INFLECT", "consensus", "consensus")
rt$engine <- factor(rt$engine,
                    levels = c("fastINFLECT 1.0", "Original INFLECT", "consensus"))

ggplot(rt, aes(x = seconds, y = method, fill = engine)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = fmt_secs(seconds)), hjust = -0.1, size = 3.6) +
  scale_fill_manual(values = c("fastINFLECT 1.0" = "#2E8B57",
                               "Original INFLECT" = "#4C78A8",
                               "consensus" = "#E07B39")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "seconds (lower is better)", y = NULL,
       title = "Scanning k = 5:25") +
  theme_minimal(base_size = 11)
```

fastINFLECT scans the whole range in **`r fmt_secs(cache$totals$inflect_scan_seconds)`**.
The original INFLECT loop is projected at
**`r fmt_secs(cache$totals$legacy_inflect_scan_seconds)`**
(**`r fmt_x(cache$totals$speedup_legacy_inflect)`** slower).
The fair consensus baseline — one ConsensusClusterPlus run — takes
**`r fmt_secs(cache$totals$amortized_seconds)`**
(**`r fmt_x(cache$totals$speedup_amortized)`** slower), and the common
per-k convenience pattern takes **`r fmt_secs(cache$totals$consensus_scan_seconds)`**
(**`r fmt_x(cache$totals$speedup)`** slower). fastINFLECT is the cheapest and
is the only one that also returns a recommended k.

---

## Why the new engine is faster

The old scan repeats work whenever the same SOM-node subtree appears in different
cuts of the hierarchy. fastINFLECT 1.0 evaluates each distinct subtree once, and then
reassembles the per-k QC matrices from that cache.

```{r efficiency-table}
eff <- cache$efficiency
eff_tbl <- data.frame(
  Quantity = c("events", "SOM nodes", "markers",
               "original cluster-marker tests",
               "memoised subtree-marker tests",
               "repeated work removed"),
  Value = c(format(eff$n_events, big.mark = ","),
            eff$n_nodes,
            eff$n_markers,
            format(eff$total_cluster_marker_tests_legacy, big.mark = ","),
            format(eff$distinct_subtree_marker_tests_inflect, big.mark = ","),
            fmt_x(eff$memoization_work_reduction)),
  check.names = FALSE
)
knitr::kable(eff_tbl, align = c("l", "r"))
```

This work reduction is separate from the faster cell kernel: the original path
uses `diptest::dip.test()` in every cluster-marker cell, while the new path uses
`diptest::dip()` plus legacy-matched p-value interpolation and small Rcpp
accelerators. The default score remains matched to the legacy score; the speed
gain comes from avoiding repeated cells and cheaper equivalent cell evaluation.

---

## Scaling with sweep density

fastINFLECT memoises QC per distinct SOM-node subtree -- and a nested (hierarchical)
sweep has at most `2·nNodes` distinct subtrees regardless of how many k you test --
so its cost is nearly flat as the sweep gets denser. The per-k consensus pattern,
by contrast, pays for another ConsensusClusterPlus run at every added k.

```{r scaling, fig.cap = "Total scan time versus the number of tested k. fastINFLECT is measured; the per-k consensus line is an optimistic lower bound (n_k times the median per-run cost measured on 5:25 -- real cost is higher because each run's cost grows with k)."}
sc <- rbind(
  data.frame(n_k = cache$scaling_df$n_k, seconds = cache$scaling_df$inflect_seconds,
             method = "fastINFLECT (measured)"),
  data.frame(n_k = cache$scaling_df$n_k, seconds = cache$scaling_df$consensus_seconds_projected,
             method = "consensus per-k (lower bound)")
)

ggplot(sc, aes(x = n_k, y = seconds, colour = method)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("fastINFLECT (measured)" = "#2E8B57",
                                 "consensus per-k (lower bound)" = "#E07B39"),
                      name = NULL) +
  labs(x = "number of k values scanned", y = "total seconds",
       title = "Cost of a denser sweep") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Because dense sweeps are essentially free for fastINFLECT, the unimodality curve -- and
the knee/threshold read off it -- can be resolved at every integer k rather than a
sparse subset. (A single ConsensusClusterPlus run is also density-independent, but
each such run costs far more than an entire fastINFLECT sweep.)

---

## Cluster quality and the recommended k

Both methods can be scored on the same objective metric: the percentage of
(cluster, marker) pairs that are unimodal (fastINFLECT's QC). The vertical guides mark
fastINFLECT's three automatic recommendations.

```{r quality, fig.cap = "Unimodality (%) across k for each method. Dashed guides mark fastINFLECT's inflection, kneedle, and unimodality-threshold recommendations."}
sel <- cache$selection
guides <- sel[is.finite(sel$k), ]

ggplot(cache$comparison_df, aes(x = k, y = unimodality, colour = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.7) +
  geom_vline(data = guides,
             aes(xintercept = k, linetype = method),
             colour = "grey30", show.legend = TRUE) +
  scale_colour_manual(values = c("fastINFLECT (hierarchical cut)" = "#2E8B57",
                                 "FlowSOM consensus" = "#E07B39"),
                      name = "scan method") +
  scale_linetype_manual(values = c(inflection = "dotted", kneedle = "dashed",
                                    threshold = "longdash"),
                        name = "fastINFLECT recommends k =",
                        labels = paste0(guides$method, " (", round(guides$k), ")")) +
  scale_x_continuous(breaks = cache$k_range) +
  labs(x = "number of metaclusters (k)", y = "unimodal (cluster, marker) pairs (%)",
       title = "Cluster quality by k") +
  theme_bw(base_size = 11) +
  theme(legend.position = "right", panel.grid.minor = element_blank())
```

```{r summary-table}
sel <- cache$selection
uni_at <- function(k) {
  d <- cache$comparison_df
  d <- d[d$method == "fastINFLECT (hierarchical cut)", ]
  hit <- d$unimodality[match(round(k), d$k)]
  if (length(hit) == 0 || is.na(hit)) NA_real_ else round(hit, 1)
}
tbl <- data.frame(
  Criterion = sel$method,
  `Recommended k` = ifelse(is.na(sel$k), "—", as.character(round(sel$k))),
  `Unimodality (%)` = vapply(sel$k, uni_at, numeric(1)),
  check.names = FALSE
)
knitr::kable(tbl, align = c("l", "r", "r"))
```

The **threshold** criterion is the most directly interpretable for the goal of
"admit no cluster with residual bimodal marker expression, without over-clustering":
it is the smallest k at which the target fraction (here
`r round(100 * ifelse(is.na(sel$target[sel$method=="threshold"]), 0.95, sel$target[sel$method=="threshold"]/ifelse(sel$target[sel$method=="threshold"]>1,100,1)))`%)
of (cluster, marker) pairs are unimodal, and it is stable across the width of the
sweep. The LL.4 **inflection** point, by contrast, depends on the fitted range.

---

## Marker-expression histograms

The aggregate unimodality score is useful, but the package promise is local:
clusters should not retain residual bimodal marker expression. The cache therefore
selects a concrete example at the threshold k. It finds a FlowSOM consensus
cluster-marker pair that fails the QC and shows overlapping fastINFLECT clusters for
the same marker that pass.

```{r marker-histogram, fig.height = 6.4, fig.cap = "Marker-expression histograms at fastINFLECT's threshold k. The selected FlowSOM consensus cluster retains residual bimodality for the marker shown; overlapping fastINFLECT clusters pass the same unimodality/spread QC."}
hist_df <- cache$marker_histogram
hist_df$qc_label <- ifelse(hist_df$qc_pass, "passes QC", "fails QC")
hist_df$panel <- factor(hist_df$panel, levels = unique(hist_df$panel))

ggplot(hist_df, aes(x = expression, fill = qc_label)) +
  geom_histogram(bins = 45, colour = "white", linewidth = 0.15) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("passes QC" = "#2E8B57", "fails QC" = "#E07B39"),
                    name = NULL) +
  labs(x = "marker expression", y = "event count",
       title = paste0("Residual bimodality example for ", unique(hist_df$marker))) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

```{r marker-table}
mh <- cache$marker_histogram_summary
mh_tbl <- data.frame(
  Method = mh$method,
  Cluster = mh$cluster,
  Marker = mh$marker,
  `QC result` = ifelse(mh$qc_pass, "pass", "fail"),
  `dip p-value` = signif(mh$dip_pvalue, 3),
  IQR = round(mh$iqr, 2),
  `events plotted` = mh$n_events,
  check.names = FALSE
)
knitr::kable(mh_tbl, align = c("l", "r", "l", "l", "r", "r", "r"))
```

In this selected example, the standard consensus cluster fails the marker
unimodality check, while the overlapping fastINFLECT clusters pass with high dip-test
p-values and narrower marker spread. This is the visual counterpart of the
quality curve: fastINFLECT's chosen k improves the actual marker distributions inside
clusters, not only the summary score.

---

## Conclusion

For the practical task of scanning k to choose a metaclustering, fastINFLECT is both
**much faster** — about `r fmt_x(cache$totals$speedup_legacy_inflect)` faster than
the original INFLECT loop in this cache, and `r fmt_x(cache$totals$speedup)` faster
than the common per-k consensus pattern — and **more decisive**, returning an
objective recommended k on the same unimodality metric used to judge every method.
Consensus metaclustering remains a reasonable way to *build* a partition at a
chosen k, but fastINFLECT is the cheaper and more informative way to *choose* that k.

---

```{r session-info}
knitr::kable(vignette_reproducibility(), align = c("l", "l"))
```
