---
title: "Benchmarking fastINFLECT against consensus metaclustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking fastINFLECT against consensus metaclustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  out.width = "100%"
)

local_cache_candidates <- c(
  file.path("..", "inst", "extdata", "benchmark_cache.rds"),
  file.path("inst", "extdata", "benchmark_cache.rds")
)
local_cache <- local_cache_candidates[file.exists(local_cache_candidates)]
cache_path <- if (length(local_cache)) {
  local_cache[[1L]]
} else {
  system.file("extdata", "benchmark_cache.rds", package = "fastINFLECT")
}
cache <- readRDS(cache_path)
library(ggplot2)
```

This vignette reports a historical engineering benchmark: the cost of scanning
a range of cluster counts on the bundled Levine32 SOM. It does not validate
cluster modality and must not be used to choose a scientific partition.

The packaged cache was generated from an early development snapshot that
reported version 1.0.0 and used `zeroes.in = FALSE`. The released interface
defaults to `zeroes.in = TRUE` and names the aggregate `qc_pass_rate`. The
cached timing remains useful as performance context. It does not support the
snapshot's aggregate “unimodality” interpretation.

## Work being compared

The historical benchmark scanned `k = 5:25` in three ways:

| Engine | Work |
|---|---|
| Original INFLECT loop | Repeats cluster-marker scoring independently at every k |
| fastINFLECT memoised Ward sweep | Scores each distinct hierarchy subtree once |
| FlowSOM consensus | Runs ConsensusClusterPlus, either once to the maximum k or repeatedly per k |

Consensus metaclustering and Ward.D2 cuts do not produce the same partitions.
The timing comparison measures the cost of scanning; it is not a claim of
methodological equivalence.

## Historical runtime

```{r runtime, fig.cap = "Historical pre-release fastINFLECT engineering timings for scanning k = 5:25 on the bundled Levine32 SOM. These bars are not modality evidence."}
runtime <- data.frame(
  method = c(
    "fastINFLECT\n(pre-release)",
    "Original INFLECT\n(estimated)",
    "FlowSOM consensus\n(single run)",
    "FlowSOM consensus\n(per-k calls)"
  ),
  seconds = c(
    cache$totals$inflect_scan_seconds,
    cache$totals$legacy_inflect_scan_seconds,
    cache$totals$amortized_seconds,
    cache$totals$consensus_scan_seconds
  ),
  stringsAsFactors = FALSE
)
runtime$method <- factor(
  runtime$method,
  levels = runtime$method[order(runtime$seconds)]
)

ggplot(runtime, aes(x = seconds, y = method)) +
  geom_col(width = 0.65, fill = "#3D7AB5") +
  geom_text(
    aes(label = ifelse(seconds < 1, sprintf("%.2f s", seconds), sprintf("%.1f s", seconds))),
    hjust = -0.1,
    size = 3.4
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    x = "wall time (seconds; lower is better)",
    y = NULL,
    title = "Historical cost of scanning k = 5:25"
  ) +
  theme_minimal(base_size = 11)
```

The original-loop bar is projected from sampled legacy
`diptest::dip.test()` calls and carries sampling error in the cache. Absolute
times are machine-specific. The structural reason for the speedup is stable:
a nested hierarchy contains far fewer distinct SOM-node subtrees than the
number obtained by summing clusters across every tested k.

```{r efficiency}
efficiency <- cache$efficiency
knitr::kable(
  data.frame(
    quantity = c(
      "events",
      "SOM nodes",
      "markers",
      "legacy cluster-marker calls",
      "memoised subtree-marker calls"
    ),
    value = c(
      format(efficiency$n_events, big.mark = ","),
      efficiency$n_nodes,
      efficiency$n_markers,
      format(efficiency$total_cluster_marker_tests_legacy, big.mark = ","),
      format(efficiency$distinct_subtree_marker_tests_inflect, big.mark = ",")
    ),
    check.names = FALSE
  ),
  align = c("l", "r")
)
```

## Release correctness gates

Performance is subordinate to the following invariants:

- `set.i` is literal and no unscheduled k is evaluated;
- every requested partition has exactly k labels and covers every SOM node;
- dip, IQR, and combined matrices are retained separately;
- complete finite transformed distributions are the default;
- simulated p-values and optional samples are deterministic across worker
  counts and preserve the caller RNG state;
- worker errors are reported before any result matrix is assembled;
- fitted endpoints without partitions are labelled as estimates;
- interpolation on tied small-sample grids is warning-free.

The package test suite checks exact serial scoring comparisons as well as
serial/two-worker identity. These are correctness checks, not evidence that a
biological cluster is homogeneous.

## Current large-model validation

The real-model audit is separate from this historical timing cache. It stages
deterministic samples from all 39,050,953 events and runs independent dip and
ACR tests with checkpointed Slurm arrays:

```{r audit-command, eval = FALSE}
system("data-raw/submit-real-model-modality.sh")
```

The workflow writes exact combinations, failures, warnings, peak RSS, wall
time, partition metadata, node-level adjusted Rand indices, and refinement
sensitivity to `inst/benchmarks/real-model-modality/`.

For the frozen BMV model
(`md5 784b3a6048c379ca64bbe8d4c0fb063d`), the literal full-event
k = 25,...,100 sweep selects combined INFLECT inflection k = 27 and
combined/dip Kneedle k = 44. The first five-k consensus plateau begins at
k = 29; IQR and event-weighted explained dispersion put the upper spread
sensitivity at k = 55. The operational fitted-model choice is nominal k = 44.
Values from k = 27 to 29 provide lower-structure sensitivities, and k = 55
provides a spread sensitivity.

One zero-event SOM-only cluster makes nominal k = 44 equal to 43
event-populated clusters. Its final independent modality audit contains 56
detected, 118 ambiguous, 28 unresolved, and 986 no-detected cluster-marker
entries. These residuals preclude a global unimodality claim. Exact selection,
consensus, internal-validity, spread, and ACR/dip evidence is retained in
`inst/benchmarks/real-model-selection/` and
`inst/benchmarks/selected-k-modality/`.

The audit uses `detected_multimodality`, `ambiguous`,
`no_detected_multimodality`, and `unresolved_discrete`. A solution-level “no
detected multimodality” statement is permitted only when every assessed marker
in every cluster has valid, concordant evidence with no ambiguous or
unresolved entry.

The early `large-model-cap-stability-2026-07-24` and
`large-model-notebook-acceptance-2026-07-24` artifacts were generated before
these rules were adopted. They support engineering provenance, not unimodality
claims, because they used the IQR-only `spread` curve, excluded non-positive
values, and did not materialise the fitted k = 62 partition.

## Reproducibility

```{r versions}
knitr::kable(
  data.frame(
    component = c("R", "fastINFLECT", "FlowSOM", "diptest", "ConsensusClusterPlus"),
    version = c(
      cache$machine$r_version,
      cache$machine$inflect_version,
      cache$machine$flowsom_version,
      cache$machine$diptest_version,
      cache$machine$consensusclusterplus_version
    ),
    stringsAsFactors = FALSE
  ),
  align = c("l", "l")
)
```
