---
title: "Comparing metaclustering strategies with fastINFLECT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing metaclustering strategies with fastINFLECT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)

# Load precomputed cache (generated by data-raw/make-vignette-cache.R)
local_cache_candidates <- c(
  file.path("..", "inst", "extdata", "vignette_cache.rds"),
  file.path("inst", "extdata", "vignette_cache.rds")
)
local_cache <- local_cache_candidates[file.exists(local_cache_candidates)]
cache_path <- if (length(local_cache)) {
  local_cache[[1L]]
} else {
  system.file("extdata", "vignette_cache.rds", package = "fastINFLECT")
}
cache <- readRDS(cache_path)
library(ggplot2)
vignette_reproducibility <- function() {
  data.frame(
    Component = c("R", "fastINFLECT", "FlowSOM", "diptest", "ConsensusClusterPlus"),
    Version = c(
      cache$machine$r_version,
      cache$machine$inflect_version,
      cache$machine$flowsom_version,
      cache$machine$diptest_version,
      cache$machine$consensusclusterplus_version
    ),
    stringsAsFactors = FALSE
  )
}
```

## Introduction

FlowSOM reduces a high-dimensional cytometry dataset to a Self-Organising Map (SOM) of a few
hundred nodes, then *metaclusters* those nodes into a biologically interpretable number of clusters
*k*. The central challenge is choosing k: too few clusters merge distinct populations; too many
split them arbitrarily.

Three common strategies exist:

| Strategy | How k is chosen |
|---|---|
| **FlowSOM consensus** | User picks k; consensus clustering (ConsensusClusterPlus) cuts the SOM |
| **FlowSOM auto-k** | `MetaClustering()` tests k = 2…max and picks the most stable k |
| **Hierarchical (ward.D2)** | User picks k; `hclust` + `cutree` cuts the SOM |
| **fastINFLECT** | Sweeps k; fits a unimodality curve; picks the inflection point automatically |

fastINFLECT provides an **objective, marker-level quality score**: for each metaclustering it measures
what fraction of (cluster, marker) combinations follow a unimodal distribution, the assumption
underlying cluster identity. As k increases, unimodality rises then plateaus; the inflection point
of that curve is fastINFLECT's recommendation.

This vignette applies all four strategies to the same SOM and compares their results using this
common metric.

---

## The data

The package ships a downsampled version of the Levine *et al.* 2015 CyTOF dataset (Levine32).
It contains 32,288 bone-marrow cells measured across 32 surface markers and pre-processed into a
FlowSOM object with **375 SOM nodes**.

```{r load-data, eval = FALSE}
library(fastINFLECT)

# Load the bundled FlowSOM object
data_path <- system.file("extdata", "Levine32sample.Rdata", package = "fastINFLECT")
load(data_path)   # creates object 'dataset' (class FlowSOM)

# 32,288 cells × 39 columns; 375 SOM nodes; 32 clustering markers
dim(dataset$data)
dataset$map$nNodes
length(dataset$map$colsUsed)
```

---

## Four ways to cut the SOM

All four approaches operate on `dataset$map$codes`, the 375 × 32 matrix of SOM node prototypes.
They each return an integer vector of length 375 assigning every node to a metacluster.

### 1. FlowSOM consensus metaclustering (fixed k)

FlowSOM's default metaclustering uses ConsensusClusterPlus to build a stable partitioning at a
user-supplied k. Here we use k = 10 as a representative choice:

```{r consensus, eval = FALSE}
library(FlowSOM)

mc_consensus <- metaClustering_consensus(
  data = dataset$map$codes,
  k    = 10,
  seed = 42
)
```

### 2. FlowSOM auto-k metaclustering

`MetaClustering()` sweeps k from 2 to a maximum and selects the most internally stable solution:

```{r autok, eval = FALSE}
mc_autok <- MetaClustering(
  data   = dataset$map$codes,
  method = "metaClustering_consensus",
  max    = 25,
  seed   = 42
)
message("Auto-k chose: ", length(unique(mc_autok)), " metaclusters")
```

### 3. Hierarchical clustering, ward.D2 (fixed k)

fastINFLECT's exported `metaClusteringhclust()` wraps `hclust(dist(..., "minkowski"), "ward.D2")` and
`cutree` at k. This is also the engine that fastINFLECT uses internally at each sweep point:

```{r hclust, eval = FALSE}
mc_hclust <- metaClusteringhclust(
  data  = dataset$map$codes,
  nClus = 10
)
```

### 4. fastINFLECT: automatic inflection-point selection

`INFLECT()` sweeps `set.i`, scores unimodality at each k via `FlowSOMQC`, fits a four-parameter
log-logistic curve, and returns the knee of that curve as its recommended k:

```{r inflect-run, eval = FALSE}
inflect_res <- INFLECT(
  FlowSOM.results = dataset,
  set.i           = 5:25,       # explicit vector, not length-2; see ?INFLECT
  multicore       = FALSE,
  zeroes.in       = FALSE
)
```

The results below come from this pre-computed run.

---

## fastINFLECT diagnostic curve

The diagnostic curve shows how unimodality improves as k increases. The vertical dashed line marks
the inflection point, where the curve stops rising, and gives fastINFLECT's recommended k.

```{r inflect-plot, fig.cap = "fastINFLECT diagnostic curve. Each point is the unimodality score at that k; the red curve is the fitted LL.4 model; the dashed vertical line is the inflection point (knee)."}
cache$inflect$ggplot
```

```{r inflect-summary}
summary(cache$inflect)
```

The `knee` column is fastINFLECT's recommended number of metaclusters. `range` is the fitted plateau
level (maximum expected unimodality); `angle` is the sharpness of the curve bend.

---

## Head-to-head comparison

We score every method on the same unimodality metric: the percentage of (cluster, marker) pairs
that pass fastINFLECT's QC test (dip test + IQR spread). Higher is better.

```{r comparison-plot, fig.cap = "Unimodality score across k for each fixed-k strategy. The dashed vertical line marks fastINFLECT's chosen knee; the triangle shows FlowSOM auto-k's chosen k and its unimodality score."}
knee    <- cache$inflect$lfunction$knee
autok   <- cache$autok_point

ggplot(cache$comparison_df, aes(x = k, y = unimodality, colour = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_vline(xintercept = knee, linetype = "dashed", colour = "black", linewidth = 0.7) +
  annotate("text", x = knee + 0.3, y = min(cache$comparison_df$unimodality),
           label = paste0("fastINFLECT\nknee = ", knee),
           hjust = 0, vjust = 0, size = 3.2) +
  geom_point(data = autok,
             aes(x = k, y = unimodality),
             shape = 17, size = 4, colour = "#7B2D8B") +
  geom_label(data = autok,
             aes(x = k, y = unimodality,
                 label = paste0("Auto-k\nk = ", k)),
             hjust = -0.15, vjust = 0.5, size = 3, colour = "#7B2D8B",
             label.size = 0.2) +
  scale_colour_manual(
    values = c("FlowSOM consensus"     = "#E07B39",
               "Hierarchical (ward.D2)" = "#3D7AB5"),
    name = NULL
  ) +
  scale_x_continuous(breaks = cache$k_range) +
  labs(
    x = "Number of metaclusters (k)",
    y = "Unimodality score (%)",
    title = "Unimodality by metaclustering strategy"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

### Summary table

```{r summary-table}
knee <- cache$inflect$lfunction$knee

# For fixed-k methods, report their best unimodality across the sweep
# (the k that maximises unimodality — their theoretical optimum if the user chose perfectly)
best_k <- function(method_name) {
  d <- subset(cache$comparison_df, method == method_name)
  d[which.max(d$unimodality), ]
}
best_consensus <- best_k("FlowSOM consensus")
best_hclust    <- best_k("Hierarchical (ward.D2)")

# fastINFLECT score at its knee
inflect_at_knee <- subset(cache$inflect$collection.U, i == knee)

tbl <- rbind(
  data.frame(
    Method            = "FlowSOM consensus",
    `k selection`     = "manual",
    `Best k in sweep` = best_consensus$k,
    `Unimodality (%)` = round(best_consensus$unimodality, 1),
    stringsAsFactors  = FALSE, check.names = FALSE
  ),
  data.frame(
    Method            = "Hierarchical (ward.D2)",
    `k selection`     = "manual",
    `Best k in sweep` = best_hclust$k,
    `Unimodality (%)` = round(best_hclust$unimodality, 1),
    stringsAsFactors  = FALSE, check.names = FALSE
  ),
  data.frame(
    Method            = "FlowSOM auto-k",
    `k selection`     = "automatic",
    `Best k in sweep` = cache$autok_point$k,
    `Unimodality (%)` = round(cache$autok_point$unimodality, 1),
    stringsAsFactors  = FALSE, check.names = FALSE
  ),
  data.frame(
    Method            = "fastINFLECT",
    `k selection`     = "automatic (inflection point)",
    `Best k in sweep` = knee,
    `Unimodality (%)` = round(inflect_at_knee$Unimodality, 1),
    stringsAsFactors  = FALSE, check.names = FALSE
  )
)

knitr::kable(tbl, align = c("l", "l", "r", "r"))
```

For fixed-k methods the table reports their **best** unimodality across the entire sweep, the
score they would achieve if a user happened to choose the optimal k. Automatic methods are shown
at their own selected k. Even under the best-case assumption for manual selection, the automatic
methods (especially fastINFLECT) reach competitive or superior scores without requiring the user to
know the right k in advance.

---

## All-cluster quality histograms

A global score can hide whether a method leaves a few poor metaclusters behind.
The plot below treats each generated metacluster as one observation and asks:
what percentage of evaluated markers pass the unimodality/IQR QC inside that
cluster? The FlowSOM consensus and hierarchical examples use the representative
fixed `k = 10` shown above; FlowSOM auto-k uses its selected k; fastINFLECT is
shown at both the original inflection criterion and the threshold criterion.

```{r cluster-quality-hist, fig.height = 7, fig.cap = "Distribution of per-metacluster unimodality on the Levine32 dataset. Each bar is the fraction of metaclusters whose marker-level QC pass rate falls in that 10-point bin. Dashed lines mark the median cluster-level score for each method."}
cq_bins <- cache$cluster_quality_bins
cq <- cache$cluster_quality_df
cq_summary <- cache$cluster_quality_summary
method_levels <- unique(cq$method)
cq_bins$method <- factor(cq_bins$method, levels = method_levels)
cq_summary$method <- factor(cq_summary$method, levels = method_levels)
cq_summary <- cq_summary[match(method_levels, as.character(cq_summary$method)), , drop = FALSE]

ggplot(cq_bins, aes(x = quality_midpoint, y = cluster_fraction)) +
  geom_col(width = 9, fill = "#3D7AB5", colour = "white", linewidth = 0.2) +
  geom_vline(data = cq_summary,
             aes(xintercept = median_cluster_unimodality),
             linetype = "dashed", colour = "#2E2E2E", linewidth = 0.5) +
  facet_wrap(~ method, ncol = 1) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20),
                     expand = expansion(mult = c(0, 0.01))) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%"),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(
    x = "markers passing QC within a metacluster (%)",
    y = "metaclusters within method (%)",
    title = "Cluster-level unimodality distributions"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r cluster-quality-table}
cq_tbl <- data.frame(
  Method = cq_summary$method,
  `k used` = cq_summary$k,
  `metaclusters` = cq_summary$clusters,
  `median cluster QC (%)` = round(cq_summary$median_cluster_unimodality, 1),
  `minimum cluster QC (%)` = round(cq_summary$min_cluster_unimodality, 1),
  `clusters below 95%` = cq_summary$clusters_below_95,
  `clusters below 100%` = cq_summary$clusters_below_100,
  check.names = FALSE
)
knitr::kable(cq_tbl, align = c("l", "r", "r", "r", "r", "r", "r"))
```

This simple view is the direct cluster-quality audit: methods whose histograms
shift toward 100% produce metaclusters where marker expression is more often
unimodal across the full marker panel. The threshold readout is included because
it is the fastINFLECT criterion that most directly encodes "avoid residual
bimodal marker expression without over-clustering."

---

## Per-marker performance

Which markers are hardest to resolve? `marker.performance()` shows, for each marker, the fraction
of metaclusters where that marker passes QC across the full sweep.

```{r marker-performance, fig.height = 5.5, fig.cap = "Per-marker unimodality pass rate (%) across all tested metaclusterings. Points are individual metaclusterings coloured by k; boxes summarise the distribution. Markers with low pass rates are systematically multimodal and may warrant panel or gating review."}
cache$mp_plot
```

Markers that consistently score low across all k values are intrinsically multimodal in this
dataset; they may reflect biphasic biology (e.g. dim vs. bright populations) that the SOM has not
fully resolved, or technical artefacts worth revisiting.

---

## Conclusion

fastINFLECT automates the most subjective step in a FlowSOM workflow, choosing k, by fitting an
objective quality curve to a unimodality score computed from the data itself. Compared with manual
selection or rule-based auto-k approaches, the inflection-point criterion is:

- **Interpretable**: the diagnostic curve directly shows where cluster quality stops improving.
- **Objective**: no heuristics or stability measures; it quantifies the actual distributional
  quality of each marker within each metacluster.
- **Reproducible**: the same input SOM always produces the same recommended k.

For downstream analysis, call `plot(inflect_res)` for the diagnostic curve, `summary(inflect_res)`
for a one-row results table, and `marker.performance(inflect_res)` to identify markers that may
require targeted gating.

---

```{r session-info}
knitr::kable(vignette_reproducibility(), align = c("l", "l"))
```
