---
title: "Use and interpret fastINFLECT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use and interpret fastINFLECT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
library(fastINFLECT)
```

fastINFLECT scans the cluster counts you request, scores marker distributions
within each partition, and reports candidate values of `k`. The candidate
values guide inspection. The retained cluster-by-marker evidence supports the
final interpretation.

## Run an analysis

To obtain these candidate values and retained evidence, start with a completed
FlowSOM or kohonen SOM. The example below uses the downsampled Levine32 FlowSOM
object included with fastINFLECT.

```{r run, message = FALSE, warning = FALSE}
load(system.file(
  "extdata",
  "Levine32sample.Rdata",
  package = "fastINFLECT"
))

schedule <- 5:12

result <- INFLECT(
  FlowSOM.results = dataset,
  set.i = schedule,
  uniform.test = "both",
  zeroes.in = TRUE,
  multicore = FALSE
)
```

`set.i` is literal. The call above evaluates every integer from 5 through 12
and stores a partition for each one. Supply at least five unique, increasing
integers, with no value above the number of SOM nodes. Use
`inflect_adaptive_set_i()` when you want a denser low-k schedule with a fixed
upper bound.

| Argument | Use |
|---|---|
| `uniform.test = "both"` | Build the aggregate from pairs that pass both dip and IQR criteria |
| `uniform.test = "unimodality"` | Build the aggregate from the dip criterion |
| `uniform.test = "spread"` | Build the aggregate from the IQR criterion |
| `zeroes.in = TRUE` | Retain all finite transformed values |
| `only.clustering.markers = FALSE` | Evaluate the names supplied in `acquired_markers` |
| `multicore = TRUE, cores = 2L` | Use two workers on platforms with fork support |

For event sampling, the default analysis uses all events. Treat
`max.events.per.node` and
`max.n.diptest` as sensitivity-analysis controls. Compare several caps and
seeds before using a sampled result.

## Read candidate values

After the analysis completes, print the result for a compact view of the run
and its candidate values. The plot shows how the selected QC pass rate changes
across the tested schedule.

```{r overview}
result
plot(result)
```

The tested scores and candidate table are available as data frames.

```{r tables}
as.data.frame(result)

result$selection[c(
  "method",
  "k",
  "qc_pass_rate_at_k",
  "partition_available",
  "k_status"
)]
```

| Method | Interpretation |
|---|---|
| `inflection` | Bend in the fitted diagnostic curve |
| `kneedle` | Largest observed departure from a straight-line increase |
| `threshold` | Smallest tested `k` that reaches `target` |

Candidate rows are diagnostic summaries, not a ranking. The example below uses
the directly tested inflection candidate only to demonstrate inspection; it is
not a default recommendation. A row with `partition_available = TRUE` was
directly tested and has an entry in `metaclustering.list`. If a finite candidate
was not directly tested, inspect `k_status`, add its `k` to a new `set.i`
schedule, and rerun before inspection. Candidate methods that happen to return
the same `k` are not independent confirmation.

## Inspect marker-level evidence

Name the candidate method, require one matching directly tested row, and use
its character value as the list key. Each pass matrix has one row per
metacluster and one column per marker.

```{r evidence}
candidate_method <- "inflection"
candidate <- result$selection[
  result$selection$method == candidate_method,
  ,
  drop = FALSE
]

if (nrow(candidate) != 1L) {
  stop("Expected exactly one inflection candidate.", call. = FALSE)
}
if (!isTRUE(candidate$partition_available[[1L]])) {
  stop(
    "Inspect `k_status`; if the candidate k is finite, add it to `set.i` and rerun.",
    call. = FALSE
  )
}

candidate_k <- candidate$k[[1L]]
key <- as.character(candidate_k)
if (!key %in% names(result$metaclustering.list)) {
  stop("The directly tested candidate partition is missing.", call. = FALSE)
}

candidate[c("method", "k", "qc_pass_rate_at_k", "partition_available", "k_status")]

partition <- result$metaclustering.list[[key]]
partition_map <- data.frame(
  som_node = seq_along(partition),
  metacluster = unname(partition)
)
head(partition_map)
table(partition_map$metacluster)

criterion_summary <- result$provenance$criterion_summary
selected_criterion_summary <- criterion_summary[
  criterion_summary$k == candidate_k & criterion_summary$selected,
  c("k", "criterion", "passed", "failed", "unresolved", "total", "qc_pass_rate"),
  drop = FALSE
]
if (nrow(selected_criterion_summary) != 1L) {
  stop("Expected exactly one selected criterion summary.", call. = FALSE)
}
selected_criterion_summary

result$criterion_pass[[key]][1:5, 1:5]
result$dip_pass[[key]][1:5, 1:5]
result$iqr_pass[[key]][1:5, 1:5]
result$combined_pass[[key]][1:5, 1:5]
```

This is an inspected candidate partition, not a scientifically selected
partition. Positions are SOM nodes and values are nominal metacluster labels.
The table counts SOM nodes, not events or cells. Labels are not ordered scores
and must not be compared numerically across different values of `k`.

`criterion_pass[[key]]` is the matrix used to calculate `qc_pass_rate`. It
equals `dip_pass[[key]]` for `uniform.test = "unimodality"`, `iqr_pass[[key]]`
for `"spread"`, and `combined_pass[[key]]` for `"both"`. `TRUE` passed the
recorded threshold, `FALSE` did not, and `NA` means that the selected decision
was unavailable. Under `"both"`, a definitive component failure can determine
`FALSE` even if the other component is unresolved, so inspect the component
matrices and `failure_reason` as well.

`qc_pass_rate` is `100 * passed / total` over the metacluster-by-marker matrix
selected by `uniform.test`; unresolved (`NA`) decisions remain in `total`.
Because `k`, cluster event counts, and test behavior change across candidates,
compare `passed`, `failed`, `unresolved`, and `total`. A higher rate means only
that a larger fraction passed. It does not imply fewer absolute failures or
that a partition is preferable or biologically valid.

Interpret the screening evidence within these limits:

- Dip-test non-rejection means no detected multimodality at the recorded
  threshold, not proof of unimodality.
- IQR is a spread screen, not a unimodality test.
- Per-pair thresholds are not multiplicity-controlled inference.
- Increasing `k` changes group sizes and test behavior, so a rising curve is
  diagnostic rather than evidence that larger `k` is biologically better.

```{r details}
details <- result$qc.details[[key]]
details$dip_p_value[1:5, 1:5]
details$iqr[1:5, 1:5]
table(details$failure_reason, useNA = "ifany")
```

Use marker summaries next to check whether a small set of markers drives the
aggregate curve.

```{r marker-performance}
marker_qc <- marker.performance(result)
head(marker_qc$marker.dataframe)
marker_qc$plot
```

## Report the result

After inspecting the marker-level evidence, record the settings that define the
run before reporting a candidate. The provenance includes the schedule, marker
panel, criterion, thresholds, value handling, sampling, seed, and worker
backend.

```{r provenance}
result$provenance[c(
  "set.i",
  "markers",
  "uniform.test",
  "th.pvalue",
  "th.IQR",
  "zeroes.in",
  "max.events.per.node",
  "seed"
)]
```

Use language that matches the evidence:

| Result | Report as |
|---|---|
| High `qc_pass_rate` | A larger fraction passed the selected QC criterion; report the counts |
| `dip_pass = FALSE` | The dip test detected evidence against unimodality at the recorded threshold |
| `iqr_pass = FALSE` | The marker IQR met or exceeded the recorded spread threshold |
| `dip_pass = TRUE` | No multimodality was detected by this dip test |
| Candidate `k` | A diagnostic candidate that still requires stability and biological review |

A criterion pass is a screening result. It does not prove unimodality,
multivariate cluster homogeneity, or biological validity. Report the selected
criterion and thresholds, confirm that the partition was directly tested, and
describe marker-level failures or unresolved measurements alongside the chosen
`k`.
