---
title: "Using fastINFLECT 1.0.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using fastINFLECT 1.0.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fastINFLECT)
```

fastINFLECT 1.0.0 uses literal cluster-count schedules, retains complete
transformed marker distributions by default, and reports dip and IQR evidence
separately. It also distinguishes fitted estimates from partitions that were
materialised.

The aggregate diagnostic remains useful for screening candidate values of
`k`. It does not establish that a cluster is truly unimodal.

## Supply a literal schedule

Candidate values enter through the required `set.i` argument. It must contain
at least five unique, strictly increasing integer values, and no value may
exceed the number of SOM nodes.

```{r literal, eval = FALSE}
schedule <- seq.int(25L, 100L, by = 5L)
print(schedule)
result <- INFLECT(model, set.i = schedule)
```

The call evaluates exactly `25, 30, ..., 100`. A two-value input such as
`set.i = c(25, 100)` fails because its spacing is ambiguous.

For an adaptive schedule with a fixed upper bound, construct the values
explicitly:

```{r adaptive}
inflect_adaptive_set_i(n_nodes = 900L, max_k = 100L)
```

`max_k` is a hard upper bound. The helper performs overflow-safe arithmetic and
refuses schedules that are unsafe to materialise.

## Retain complete transformed values

With the schedule defined, `zeroes.in = TRUE` retains negative, zero, and
positive finite values. This matters for centred or otherwise transformed
marker data because dropping non-positive measurements changes the tested
distribution.

```{r complete-values, eval = FALSE}
result <- INFLECT(
  model,
  set.i = schedule,
  zeroes.in = TRUE
)
```

Conversely, `zeroes.in = FALSE` excludes every non-positive value, warns when
negative inputs are present, and records the exclusions:

```{r excluded-values, eval = FALSE}
compatibility_result <- INFLECT(
  model,
  set.i = schedule,
  zeroes.in = FALSE
)

compatibility_result$provenance$zero_handling
```

Under either `zeroes.in` setting, no zero-valued padding observations are
manufactured.

## Read separated criteria

`uniform.test` selects the aggregate pass-rate curve. All component evidence
remains available:

| `uniform.test` | Aggregate curve | Evidence still retained |
|---|---|---|
| `"both"` | dip and IQR pass | dip, IQR, combined |
| `"unimodality"` | dip pass | dip, IQR, combined |
| `"spread"` | IQR pass | dip, IQR, combined |

```{r criterion-evidence, eval = FALSE}
result <- INFLECT(
  model,
  set.i = schedule,
  uniform.test = "both"
)

result$scores
result$dip_pass[["50"]]
result$iqr_pass[["50"]]
result$combined_pass[["50"]]

details <- result$qc.details[["50"]]
details$dip_p_value
details$iqr
details$event_count
details$test_event_count
details$failure_reason
```

The canonical aggregate column is `qc_pass_rate`. Compatibility aliases from
the original INFLECT interface remain available but are deprecated:

| Deprecated | Canonical |
|---|---|
| `U.set`, `collection.U` | `scores` |
| `Unimodality` | `qc_pass_rate` |
| `Accuracy.sets`, `Accuracy.matrixes` | `criterion_pass` |

## Opt in to parallel scoring

With the result fields defined, serial scoring is the default. Parallel
scoring requires at least two workers:

```{r parallel, eval = FALSE}
result <- INFLECT(
  model,
  set.i = schedule,
  multicore = TRUE,
  cores = 2L
)
```

Unix-like systems use forked workers over distinct SOM-node subtrees.
Unsupported platforms validate the same inputs and use a recorded serial
fallback. Every worker result is checked for schema, marker order, and length
before matrices are assembled.

Simulated dip p-values and optional event sampling use deterministic,
overflow-safe streams derived from the subtree and marker. They preserve the
caller's RNG state and return the same result with one or two workers.

## Use event caps for sensitivity analysis

`max.events.per.node` samples each node once with `seed`. The same retained
rows feed every marker, both criteria, every k, and every worker:

```{r sampling, eval = FALSE}
result <- INFLECT(
  model,
  set.i = schedule,
  max.events.per.node = 2000L,
  seed = 42L
)

result$provenance$scoring_mode
result$provenance$n_events_retained
result$provenance$event_sampling$per_node
```

The default `NULL` retains every event. A cap should only be adopted after
checking score and endpoint stability across relevant sample sizes and seeds.

## Distinguish estimates from partitions

With the event-sampling choice established, `result$selection` reports
`partition_available`, `k_status`, and `score_source`:

```{r selection, eval = FALSE}
result$selection[c(
  "method",
  "k",
  "qc_pass_rate_at_k",
  "partition_available",
  "k_status",
  "score_source"
)]
```

A fitted-curve inflection can be an unscheduled integer. In that case it is
labelled `fitted_estimate_no_partition`; fastINFLECT does not silently
evaluate it. To obtain the partition and its criterion evidence, include that
literal k in a new schedule.

No value of k should be promoted solely because it sits at the inflection of an
aggregate pass-rate curve.

## Audit provenance before interpretation

The QC boundary records the criterion, thresholds, marker set, zero handling,
sampling order, seeds, worker backend, model layer weights, retained events,
and per-criterion summaries:

```{r provenance, eval = FALSE}
str(result$provenance, max.level = 2)
```

Then use the following language for downstream validation:

- `detected_multimodality` when independent tests reproducibly reject;
- `no_detected_multimodality` when valid tests do not reject;
- `ambiguous` for test disagreement, sensitivity, or borderline evidence;
- `unresolved_discrete` for insufficient unique values, excessive ties,
  saturation, or test failure.

Of these, “no detected multimodality” is deliberately weaker than “unimodal.”
The dip test can reject a unimodal null; failure to reject does not prove that
null.
